---
title: "Course Project 1"
author: "M Dupont"
date: "23 October 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(ggplot2)
```

## R Markdown

Import .csv file containing data from working directory

```{r}
data <- read.csv("activity.csv")

```

## What is the mean total number of steps taken per day? 

 Convert date column to date format and sum total steps for each day
```{r}
data$date <- as.Date(as.character(data$date))
steps.day <- as.data.frame(data %>% group_by(date) %>% summarise(total.steps=sum(steps)))
```

Make a histogram of the total steps taken each day
```{r figure_1}
hist(steps.day$total.steps,breaks=12, col="blue",main="Total Steps per Day",xlab="Total Daily Steps")
```


Calculate the mean and median of the total number of steps taken each day
```{r}
mean.steps <- format(mean(steps.day$total.steps,na.rm=TRUE), digits=7)
print(mean.steps)
median.steps <- format(median(steps.day$total.steps,na.rm=TRUE),digits=7)
print(median.steps)
```
The mean number of steps taken per day is `r mean.steps` and the median steps taken per day is `r median.steps`. 

## What is the average daily activity pattern? 
Group data by interval and calculate average steps for each interval
```{r figure_2}
steps.int <- as.data.frame(data %>% na.omit %>% group_by(interval) %>% summarise(av.steps=mean(steps)))

par(mar=c(5,4,1,1),las=1)
plot(steps.int$interval,steps.int$av.steps,type='l',main="Average Steps per Interval",xlab="interval",ylab="average steps")
```

Find the interval corresponding to the maximum average steps
```{r}
max.int <- steps.int$interval[which.max(steps.int$av.steps)]
print(max.int)
```

The 5 minute interval which contains the maximum number of average steps is `r max.int`. 

## Input missing values

Calculate the total number of missing values (NAs)
```{r}
tot.NA <- sum(is.na(data$steps))
print(tot.NA)
```

The total number of missing values (NAs) is `r tot.NA`.

Replace all NA values with the mean value for the interval over all days.
```{r}
impute.mean <- function(x) replace(x,is.na(x),mean(x,na.rm=TRUE))
na.replace <- ddply(data,~interval,transform,steps=impute.mean(steps))
data.na.rm <- as.data.frame(data.na.rm <- na.replace %>% group_by(date) %>% summarise(daily.steps=sum(steps)))
```
Create histogram of the total number of steps taken each day and calculate mean and median values.

```{r figure_3}
hist(data.na.rm$daily.steps,breaks=12,col="red",xlab="Average Steps per day", main="Average Steps per Day (NA replaced)")
mean.na.rm <- format(mean(data.na.rm$daily.steps), digits=7)
print(mean.na.rm)
median.na.rm <- format(median(data.na.rm$daily.steps), digits=7)
print(median.na.rm)
```
The mean total number of daily steps is `r mean.na.rm`, and the median is `r median.na.rm`. The mean number of steps is the same as for the dataset without NAs replaced. The median number of steps has changed from `r median.steps` to `r median.na.rm`


## Are there differences in activity patterns between weekdays and weekends?

Create factor variable designating 'weekday' or 'weekend'.
```{r}
weekdays1 <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
na.replace$wday <- factor((weekdays(na.replace$date) %in% weekdays1), levels=c(FALSE,TRUE),labels=c('weekend','weekday'))
```
Make a panel plot containing data averaged over 5 min intervals for weekdays and weekends

```{r figure_4}
data.wday <- as.data.frame(na.replace %>% group_by(interval,wday) %>% summarise(av.steps=mean(steps)))

p1 <- ggplot(data.wday, aes(x=interval,y=av.steps)) +
               geom_line()+
               facet_wrap(.~wday,nrow=2,ncol=1)+
               xlab("Interval")+
               ylab("Average Steps")

print(p1)
```

The activity pattern of average steps for each interval is different for weekdays and weekends. 
